Escaping RGBland: Selecting Colors for Statistical Graphics

### 2007-11-07 pkg/paper/Palettes/CSDA.Rnw
added Tufte ref and some more touches
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@29 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%FIXME% Different/better title?
\title{Escaping RGBland: selecting colors for \\statistical graphics}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
   \And Paul Murrell\\The University of Auckland}
\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

\textit{The colors should not be unappealing.}
It is not necessary for the colors in a statistical plot to reflect 
fashion trends, but basic principles such as avoiding large areas
of fully saturated colors \citep{Tufte-1990-EIN}
should be adhered to. 
The requirement is not that the user should have
a degree in graphic design, but that the software should provide users with an
intuitive way to select colors and control their basic properties. 
Thus, it is necessary to employ a \emph{color model}
or \emph{color space} that describes colors in terms of their perceptual
properties: hue, brightness, and colorfulness.

A color model typically supported by software packages involves
 the specification
of colors as Red-Green-Blue (RGB) triplets. However, this specification
corresponds to color generation on
a computer screen 
\citep[see][for more details]{color:Poynton:2000}
rather than
corresponding to human color perception. For humans, it is
virtually impossible to control the perceptual properties of a color in this
color space because there is no single dimension that corresponds to, for example,
the hue or the brightness of the color. As a consequence, 
various perceptually-based
color spaces have been suggested, where each dimension of the color space can
be matched with a perceptual property. One approach widely implemented in
software packages involves
 Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
a simple transformation of RGB triplets \citep[see]{color:Wiki+HSV:2007}.
Unfortunately, the dimensions in HSV space map poorly to perceptual
properties and the use of HSV colors encourages the use of 
highly saturated colors. 
A perceptually based color model that avoids
these problems involves
Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
which is
a transformation of CIELUV color space \citep[see]{color:Wiki+CIELUV:2007}.
This is the colorspace we advocate in this article.

\textit{The colors in a statistical graphic should cooperate with each other.}
The typical purpose of color in a statistical graphic is to distinguish
between different areas or symbols in the plot---to distinguish between
different groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and that those colors should be related to
each other.

A natural solution for this task is to vary (at least) one perceptual property
of the colors, e.g., the hue or the brightness, keeping other properties
fixed. In a perceptually based color space, this corresponds to selecting
colors by traversing paths through this space along its dimensions.
This approach is implemented in many color picker tools
\citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
however, most are commonly based on the HSV model.
As argued above, the dimensions in HSV space are not truly independent and
hence it is not possible to vary just one perceptual property while keeping
the others fixed. This means that it is relatively
difficult to select sets of HSV coordinates that yield colors that are
``in harmony'' \citep[see][]{color:Munsell:1905}. For statistical graphics,
this is important because it can introduce size distortions in the perception
of shaded areas and can produce
 optical illusions \citep{color:Cleveland+McGill:1983}.
These problems can again be addressed by employing an HCL colorspace. 
In addition to
the earlier point that HCL space
is useful because it allows humans to understand where within the space
a particular HCL triplet is located,  
HCL also allows
us to understand motion within the color space because
distances between colors  have an intuitive and rational meaning.

\textit{The colors should work everywhere.}
The final issue to deal with is that, in an ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distinguishable when the graphic is displayed on an LCD projector
rather than a computer screen, or when it is printed on a grayscale
printer, or when the person viewing the graphic is color-blind.
These goals cannot always be attained, but attention should be paid
to these issues and in many situations it is also possible to
resolve any problems.

An example of this approach is \pkg{ColorBrewer.org}
\citep{color:Harrower+Brewer:2003}, an online tool for selecting color
schemes for maps. It provides a collection of prefabricated palettes, 
supplies information about the
suitability for color-blind viewers or grayscale printing, and guides its
users how to choose a suitable palette for coding various types of
information. The drawback to this tool is that it only provides a fixed
set of colors for each palette and there is no way to extend the
existing palettes.

@@

To show what can be gained by selecting appropriate color schemes,
we present two examples for typical color graphics, contrasting
commonly-used HSV palettes with more suitable HCL palettes.
Both examples and all HSV palettes are taken from (the electronic
version of) recent publications in
\textit{Computational Statistics \& Data Analysis} (Volume~51).

@@

In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
coding the variable of interest much better and without flashy colors. If, however, the user wants
to increase the contrast by adding some color to the plot, this could be done by using a better
balanced version of the heat colors (as shown in the bottom right panel). These colors also
increase from a yellow to a red hue while using the same brightness levels as in the grayscale
palette. Thus, when converted to a grayscale or printed out on a grayscale printer
(as in the printed version of \textit{Computational Statistics \& Data Analysis}),
the upper and lower right panel would look (virtually) identical.
Both palettes have in common that they give increasing perceptual emphasis
to regions with increasing density, resulting in a heatmap that highlights the (small)
interesting high-density regions and not to the large low-density regions surrounding them.
By balancing the lower right panel with respect to its brightness levels (i.e., light/dark contrasts),
it is assured that the graphic is intelligible for color-blind readers and that the same graphic works in
colored electronic versions and printed grayscale versions of a publication.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<wheel, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
par(lwd=3)
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
par(lwd=1, new=TRUE)
pie(rep(1,60), radius = 1, col = rainbow(60), 
    border=NA, labels = NA)
par(lwd=3)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), 
    labels = 360 * 0:5/6)
par(lwd=1, new=TRUE)
pie(rep(1,60), radius = 1, col = rainbow_hcl(60, c = 50, l = 70), 
    border=NA, labels = NA)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

>>>>

\title{Escaping RGBland: Selecting Colors for Statistical Graphics}
\author{Achim Zeileis\\Wirtschaftsuniversit\"at Wien
   \And Kurt Hornik\\Wirtschaftsuniversit\"at Wien
   \And Paul Murrell\\The University of Auckland}
\Plainauthor{Achim Zeileis, Kurt Hornik, Paul Murrell}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

\textbf{The colors should not be unappealing.}
It is not necessary for the colors in a statistical plot to reflect 
fashion trends, but basic principles such as avoiding large areas
of fully saturated colors \citep{color:Tufte:1990}
should be adhered to. 
The requirement is not that the user should have
a degree in graphic design, but that the software should provide users with an
intuitive way to select colors and control their basic properties. 
Thus, it is necessary to employ a \emph{color model}
or \emph{color space} that describes colors in terms of their perceptual
properties: hue, brightness, and colorfulness.

A color model typically supported by software packages involves
 the specification
of colors as Red-Green-Blue (RGB) triplets. However, this specification
corresponds to color generation on
a computer screen 
\citep[see][]{color:Poynton:2000}
rather than
corresponding to human color perception. For humans, it is
virtually impossible to control the perceptual properties of a color in this
color space because there is no single dimension that corresponds to, e.g.,
the hue or the brightness of the color. As a consequence, 
various perceptually-based
color spaces have been suggested, where each dimension of the color space can
be matched with a perceptual property. One approach widely implemented in
software packages involves
 Hue-Saturation-Value (HSV) triplets \citep{color:Smith:1978},
a simple transformation of RGB triplets \citep[see][]{color:Wiki+HSV:2007}.
Unfortunately, the dimensions in HSV space map poorly to perceptual
properties and the use of HSV colors encourages the use of 
highly saturated colors. 
A perceptually based color model that avoids
these problems involves
Hue-Chroma-Luminance (HCL) triplets \citep[see][]{color:Ihaka:2003},
which is
a transformation of CIELUV color space \citep[see][]{color:Wiki+CIELUV:2007}.
This is the colorspace we advocate in this article.

\textbf{The colors in a statistical graphic should cooperate with each other.}
The typical purpose of color in a statistical graphic is to distinguish
between different areas or symbols in the plot---to distinguish between
different groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and that those colors should be related to
each other.

A natural solution for this task is to vary (at least) one perceptual property
of the colors, e.g., the hue or the brightness, keeping other properties
fixed. In a perceptually based color space, this corresponds to selecting
colors by traversing paths through this space along its dimensions.
This approach is implemented in many color picker tools
\citep{color:Moretti+Lyons:2002,color:Meier+Spalter+Karelitz:2004},
however, these are typically based on the HSV model.
As argued above, the dimensions in HSV space are not truly independent and
hence it is not possible to vary just one perceptual property while keeping
the others fixed. This means that it is relatively
difficult to select sets of HSV coordinates that yield colors that are
``in harmony'' \citep[see][]{color:Munsell:1905}. For statistical graphics,
this is important because it can introduce size distortions in the perception
of shaded areas and can produce
 optical illusions \citep{color:Cleveland+McGill:1983}.
These problems can again be addressed by employing the HCL colorspace. 
In addition to
the earlier point that HCL space
is useful because it allows humans to understand where within the space
a particular HCL triplet is located,  
HCL also allows
us to understand motion within the color space because
distances between colors  have an intuitive and rational meaning.

\textbf{The colors should work everywhere.}
The final issue to deal with is that, in an ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distinguishable when the graphic is displayed on an LCD projector
rather than a computer screen, or when it is printed on a grayscale
printer, or when the person viewing the graphic is color-blind.
These goals cannot always be attained, but attention should be paid
to these issues and in many situations it is also possible to
resolve any problems.

An example of this approach is \pkg{ColorBrewer.org}
\citep{color:Harrower+Brewer:2003}, an online tool for selecting color
schemes for maps. It provides a collection of prefabricated palettes, 
information about the suitability for printers or color-blind viewers,
and guidance on how to choose a suitable palette for coding various types of
information. The drawback to this tool is that it only provides a fixed
set of colors for each palette and there is no way to extend the
existing palettes.

@@

To show what can be gained by selecting appropriate color schemes,
we present two examples for typical color graphics, contrasting
commonly-used HSV palettes with more suitable HCL palettes.
Both examples and all HSV palettes are taken from (the electronic
version of) recent publications in
\emph{Computational Statistics \& Data Analysis} (Volume~51).

@@

In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
coding the variable of interest much better and without flashy colors. If, however, the user wants
to increase the contrast by adding some color to the plot, this could be done by using a better
balanced version of the heat colors (as shown in the bottom right panel). These colors also
increase from a yellow to a red hue while using the same brightness levels as in the grayscale
palette. Thus, when converted to a grayscale or printed out on a grayscale printer
(as in the printed version of \emph{Computational Statistics \& Data Analysis}),
the upper and lower right panel would look (virtually) identical.
Both palettes have in common that they give increasing perceptual emphasis
to regions with increasing density, resulting in a heatmap that highlights the (small)
interesting high-density regions and not to the large low-density regions surrounding them.
By balancing the lower right panel with respect to its brightness levels (i.e., light/dark contrasts),
it is assured that the graphic is intelligible for color-blind readers and that the same graphic works in
colored electronic versions and printed grayscale versions of a publication.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<wheel, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
par(lwd=3)
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
par(lwd=0.5, new=TRUE)
pie(rep(1, 100), radius = 1, col = rainbow(100), border = NA, labels = NA)
par(lwd=3)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), 
    labels = 360 * 0:5/6)
par(lwd=0.5, new=TRUE)
pie(rep(1, 100), radius = 1, col = rainbow_hcl(100, c = 50, l = 70), border = NA, labels = NA)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}
